State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources

### 2020-08-18 analysis/paper/paper.Rmd
revised document and added SimFnc.R

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Mike W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 3.5
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  This is the abstract
keywords: |
  escapement; Chinook; steelhead; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

* **[North American Journal of Fisheries](https://afspubs.onlinelibrary.wiley.com/journal/15488675)**

@@

Salmon escapement is the number of adults that escape harvest and return to their freshwater habitat to potentially spawn (e.g. @Bue1998). For anadromous fishes, escapement is often estimated at a given point in a river system prior to fish reaching their spawning area. Accurate escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement by groups (e.g., stock, population, origin [hatchery, wild]) provide valuable information that fisheries managers can use to achieve sustainable harvest of larger groups, while protecting vulnerable populations. Accurate escapement estimates are increasingly important for depleted populations as they facilitate assessments of population viability and extinction risk. In this study, we describe a novel method to estimate escapement, with uncertainty, using multiple sources of data and demonstrate that the method is essentially unbiased. Moreover, the method allows escapement to be parsed by groups (e.g., wild, hatchery) which is important towards effective conservation and management decision making.

Populations of Chinook salmon Oncorhynchus tshawytscha and steelhead trout O. mykiss in the Snake River basin of Pacific Northwest, USA, have been depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats. As a result, Snake River spring-summer run Chinook salmon (hereafter Chinook salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA); Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later. Snake River Chinook salmon and steelhead have substantial recreational, commercial, and subsistence value both within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries. The aggregate escapement of Snake River Chinook salmon and steelhead populations, with the exception of Tucannon River, is measured at Lower Granite Dam located in southeast Washington, the final dam on the Snake River that returning adults must ascend prior to heading to tributary habitats to spawn. Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin (hatchery, wild). Harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia River fisheries, can be predicated on escapements at Lower Granite Dam. Moreover, escapement of wild populations is used to inform viable salmonid population (VSP) analyses used for ESA status reviews.

Chinook salmon and steelhead returning to a majority of populations in the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary habitats to spawn. Previously, management agencies used counts of fish, by species, passing an observation window located on the side of the fish ladder as a census of fish migrating past the dam (i.e., Lower Granite escapement). For both species, total escapement was then parsed to groups (wild, hatchery clipped, hatchery unclipped) using mark and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window [@Camacho2017]. Window counts as a census proved beneficial as being an easy, straight-forward method that could be ascertained in near real-time. Moreover, window count estimates have been used in downriver fisheries management arenas for several years and consistency in methods is often desirable for management decision making.

However, using window counts as a census of Chinook salmon and steelhead passing Lower Granite Dam can be problematic as it avoids multiple sources of uncertainty and ignores known biological processes. First, live window counts only occur from April through October each year. The live window counts only occur for 16 hours each day; fish counters working at the observation window look directly into the fish ladders to count fish, by species, passing by for 50 minutes of each hour. Counts are then multiplied to provide an estimate for the hour. From November through March, the remainder of the year, video tape fish counting is used and only occurs for 10 hours each day; fish counters then reach the tapes and submit daily fish counts [@Hatch1994]. Typically, the error rates of live and video window counts are unknown. Additionally, two biological processes are unaccounted for: 1) fish that cross the dam while the window is closed for counting (i.e., nighttime passage), and 2) fish that cross the dam multiple times (i.e., re-ascension) and therefore potentially double-counted. Uncounted nighttime passage potentially results in an underestimate of escapement. Finally, Chinook salmon and steelhead that migrate through the ladder and past the dam may fallback over the dam (e.g., over spillway, through navigation locks), and later, may or may not re-ascend the fish ladder, again [@Boggs2004]. Both fallback with no re-ascension and fallback with re-ascension potentially result in an overestimate of escapement; re-ascension can result in double-counting of fish [@Dauble2000]. Previously, it was assumed that nighttime passage rates and fallback/re-ascension rates cancelled each other out resulting in window counts providing an unbiased estimate of escapement; we further evaluate that assumption here.

Here, we present a novel method for estimating escapement, by species, past Lower Granite Dam that incorporates window counts, data from the adult fish trap, and observations of PIT-tagged fish in the adult ladder and that explicitly models nighttime passage, re-ascension and error from both window and trap estimates. To meet desired management and conservation objectives, total escapement estimates include estimates of uncertainty and are parsed into weekly strata. Further, weekly estimates are parsed into three origin groups: wild, hatchery, and hatchery no-clip. Estimates of escapement account for fish that migrate through the ladder at night outside of observation hours and account for fish that may ascend the ladder multiple times. Our methods is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM). The STADEM model is available as a package for the statistical software R [@R-Core-Team2020], and can be downloaded from a [GitHub site](https://github.com/KevinSee/STADEM). To validate the model, we simulate 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates. The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provide more reasonable estimates of uncertainty. This STADEM framework could be applied elsewhere to any fish passage barrier with a counting mechanism like a window, an adult trap that samples a portion of the run and contains a PIT tag antenna (or similar).

# Materials and Methods

@@

We used STADEM and three sources of data to estimate Chinook salmon and steelhead escapement at Lower Granite Dam. Data sources included 1) counts of fish migrating pas an observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of previously PIT tagged fish detected in the adult fish ladder. Following, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; however, similar data could also likely be obtained from other fish passage facilities.

@@

Daily counts of adult Chinook salmon and steelhead passing an observation window located on the Lower Granite fish ladder are the first source of data. Daily counts are made and provided by the US Army Corps of Engineers (COE) and, when summed, provide an estimate of the number of fish ascending and passing (i.e., escapement) Lower Granite Dam each season. Window counts are made for each species using both video monitoring and direct visual monitoring during daytime hours [@Hatch1994]. Video monitoring occurs during the beginning and tail ends of the adult runs (March 1 – March 31 and November 1 – December) and for 10 hours per day (0600 – 1600 hours). Direct visual monitoring occurs during peak run times (April 1 – October 31) and for 16 hours per day (0400 – 2000 hours) [@USACE2015]. During direct visual monitoring, observers record each adult (≥ 30cm), by species, passing the window for 50 minutes of each hour of operation. Salmonids under 30cm in length are not identified to species. The sum of the daily 50-minute counts are then multiplied by 1.2 to account for the 10 minutes where fish are not counted. These daytime window counts are not expanded for fish that may ascend the ladder outside of operational hours (i.e., nighttime) [@USACE2015]. Window counts are accessed through the Columbia Basin Research Data Access in Real Time (DART) [website](www.cbr.washington.edu/dart/query/adult_daily), using their window count query. Counts are provided for each day the fish ladder is open to passage. Although window counts are assumed to be a census of every fish passing Lower Granite Dam, nighttime passage and re-ascending fish are not accounted for. Further, there is no estimation of daily or seasonal observation error.

@@

The second source of data comes from a trap that captures a sample of the adults as they migrate past Lower Granite Dam [@Ogden2016b]. The trap, also located within the adult fish ladder and upstream of the observation window, provides biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) for captured adults that allows decomposition of the escapement into specific groups (e.g., @Camacho2019, @Steinhorst2017). The trap randomly samples the daily run by opening four times per hour for a length of time determined by a set daily trapping rate. The trap is operational 24 hours per day. The trap rate is determined by a committee of collaborating management agencies with a goal of capturing a target number of wild fish but also balancing fish handling concerns. Trap sample rates are typically 10-25%, but may change throughout the season due to high water temperatures, decreased flows, trap malfunctions and/or closures, fish handling logistics, etc. All captured fish are anesthetized, identified to species, examined for marks/tags, and measured for fork length (FL). Additionally, each fish has scale and genetic tissue samples taken; scale samples are used to estimate age [@Wright2015], tissue samples are used to determine sex and for genetic stock identification analysis (e.g., @Hargrove2019). Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) receive a PIT tag. Final determination of wild, hatchery, and hatchery no-clip origins are assigned using a post-hoc analysis of marks, tags, and genetic information. Data from the adult trap are collected and managed by multiple agencies and are made available by the Idaho Department of Fish and Game (IDFG) in a comma separated value (.csv) file [@Camacho2019].

### PIT-tag Data

The last source of data is observations of PIT tagged adult Chinook salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder. These observations provide estimates of 1) a trapping rate, 2) the proportion of fish passing during nighttime hours (i.e., outside of window observation hours), and 3) the proportion of fish that ascend the fish ladder multiple times (i.e., re-ascension rate). Detections used in the model include all fish that were previously PIT tagged as juveniles or adults prior to reaching Lower Granite Dam (i.e., not those tagged at the dam as adults) and detected at adult detection sites in the dam passage system. Data are provided through DART and the [adult ladder PIT tag query](http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail).

A trap rate estimate is derived using mark-recapture methods and PIT tag observations of both Chinook salmon and steelhead at Lower Granite Dam adult detection sites. The "mark" group includes all tags detected in the adult trap and the "capture" group includes tags observed to cross the weir at the upstream end of the fish ladder as adults leave the passage system. Using a mark-recapture model with differing capture probabilities, we can estimate the trap rate on a weekly basis. Those estimates, with associated uncertainties, are then incorporated into the model as informed priors, while the model estimates the true trap rate based on all the data, including trap and window counts. The true trapping rate is estimated as the recorded time of the trap being open does not always reflect the true proportion of fish that are captured in the trap, due to trap malfunctions, sort-by-code fish opening the trap more frequently than expected, or process error (Rick Orme, *per. comm.*). Therefore, we use the mark-recapture estimate of trap rate.

The nighttime passage rate is estimated simply as the proportion of tags that migrate through the fish ladder during "nighttime" hours outside of the window observation hours divided by the total number of tags passing the fish ladder and is calculated on a weekly basis. The re-ascension rate is the number of tags observed passing the upstream most detection sites in the adult fish ladder (i.e., passing the dam) and later detected re-entering the downstream end of the fish ladder at a later time divided by the total number of tags leaving the fish ladder. Previously, we examined for differences in nigh passage and re-ascension rates estimated using wild fish, versus combining hatchery and wild fish together, and found no difference. Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.

@@

STADEM estimates the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage and fallback-and-reascension rates. Using a state-space approach, STADEM assumes that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) are generated by processes with observation error. In the case of the trap, there is sampling variation and uncertainty in what the true trap rate is. STADEM further accounts for the proportion of fish that pass while the counting window is closed (i.e., night), as well as the potential double-counting (or more) of fish that have fallen back below the dam and later re-ascended the fish ladder. Next, origin data (wild, hatchery, hatchery no-clip) are used to partition the total escapement estimate by origin (Figure \@ref(fig:stadem-examp-fig). Additional model details can be found in [Appendix A](#append1). STADEM is implemented in R software [@R-Core-Team2020] and is available from the primary author at a [GitHub site](https://github.com/KevinSee/STADEM).

@@

We tested the STADEM model on a variety of simulated data sets. These simulated data sets contained a fixed number of unique adult fish of known origin crossing a dam, from a total of 25 fictional populations with differential run-timing (i.e., date of passage at Lower Granite Dam). Each simulated fish is given a date of ascending the ladder, based on its population and the range of run timing for that population. Each fish is also simulated to cross the dam either while the window is open for counting, or not, and is given the change to be "caught" in the simulated fish trap given the week when it ascends the dam, and the known trap rate that week. Fallback and re-ascension behavior was also simulated, with each fish having the possibility of falling back and re-ascending the ladder up to three times.

Our objective was to examine STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement from the combinations of two separate trapping rates, two fallback, re-ascension and nighttime passage combinations, and three window count error rates; resulting in twelve different scenarios (Table \@ref(tab:lgr-sim-tab)). The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at Lower Granite Dam between 2010-2015. Further details about simulation procedures can be found in [Appendix B](#append2).

@@

Finally, we applied STADEM to data from Lower Granite Dam for Chinook salmon and steelhead returning to the Snake River to spawn during 2010 to 2019. Window counts for both species were accessed from DART via STADEM. For spring-summer run Chinook salmon, a spawn year refers to adults that migrate pas the dam prior to August 17 each year and spawn that late summer and fall. For steelhead, the spawn year is defined as steelhead that migrate past Lower Granite Dam starting July 1 the previous year and prior to June 30 of the given year (e.g., spawn year 2017 steelhead migrate past Lower Granite between July 1, 2016 and June 30, 2017). Data from the adult trap was made available by IDFG and adult detection data at Lower Granite Dam from the fish passage ladder is available in the regional database PTAGIS ([PIT Tag Information System](https://www.ptagis.org/)).

@@

STADEM results were very similar across all scenarios (Figure \@ref(fig:rel-bias-lgd)). Estimates were unbiased, with an average relative bias of 0.2 – 0.3%. The CV of the estimates average 2 – 3%, with coverage probabilities that always exceeded 95%. We calculated RMSD as the square root of the sum of the variance of the estimate and the squared expected bias, which accounts for the size of the uncertainty in the estimator as well as its bias. The RMSD was near 500 for each scenario, representing an estimate within 2% of the true value (Table \@ref(tab:summ-tab)).

@@

We applied STADEM to data from Lower Granite Dam for Chinook salmon and steelhead for spawn years 2010 – 2019. Estimates of total escapement, as well as estimates of wild, hatchery, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab). Coefficients of variation range from 3.5 - 7.5% for wild fish, 3.0 - 6.2% for hatchery fish, 4.0 - 8.4% for hatchery no-clip fish and 2.9 - 5.6% for total unique fish past Lower Granite Dam.

>>>>

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Mike W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  This is the abstract
keywords: |
  escapement; Chinook; steelhead; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

* **[North American Journal of Fisheries Management](https://afspubs.onlinelibrary.wiley.com/journal/15488675)**

@@

Fish escapement is the number of adults that survive juvenile and adult rearing, escape harvest and return to their natal habitat to potentially spawn (e.g. @Bue1998). For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area. Escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement for each component group (e.g., stock, population, age, origin [wild, hatchery]; @hess2014monitoring; @steinhorst2017abundance) provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations. Accurate escapement estimates are increasingly important for depleted populations as they inform viable salmonid metrics [@mcelhany2000viable], and facilitate assessments of population viability and extinction risk [@nwfsc2015; @williams2016viability] and climate change vulerability [@crozier2019climate].

Populations of Chinook Salmon Oncorhynchus tshawytscha and steelhead trout O. mykiss in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats [@nehlsen1991pacific; @mcclure2003large; @nwfsc2015]. As a result, Snake River spring-summer run Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937). Snake River Chinook Salmon and steelhead have substantial recreational, commercial, cultural and subsistence value both within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries. The aggregate escapement of Snake River Chinook Salmon and steelhead populations, with the exception of Tucannon River, is measured at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations. Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin (**Snake Basin Harvest Plans?**). Additionally, harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia River fisheries, are predicated on escapements at Lower Granite Dam (**could cite US v OR Management Agreement and Snake Basin Harvest Plans?**).

Chinook Salmon and steelhead returning to a majority of populations in the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary spawning locations. Fish management agencies previously used counts of fish, by species, passing an observation window located on the side of the fish ladder as a census of fish escaping past the dam. For both species, total escapement was then parsed into groups (e.g., wild, hatchery) using observed marks and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window [@Camacho2017, @Steinhorst2017]. Window counts as a census proved beneficial as being an easy, straight-forward method that was ascertained in near real-time. Moreover, downriver fisheries management arenas have used window counts at lower-river dams as escapement estimates for the past several years, and consistency in methods is often desirable for management decision making (**citation?**).

Using window counts, however, as a census of Chinook Salmon and steelhead passing Lower Granite Dam can be problematic as it avoids multiple sources of uncertainty and ignores known biological processes. First, live (i.e., in-person) window counts only occur from April through October each year. Additionally, live window counts only occur for 16-hours each day, and fish counters working at the observation window look directly into the fish ladders to identify and count all passing fish, by species, for only 50 minutes of each hour. Counts are then expanded to provide an estimate for the hour [@USACE2015]. From November through March, the remainder of the year, video tape fish counting is used and only occurs for 10 hours each day; fish counters then read the video tapes and submit daily fish counts [@Hatch1994]. Typically, the observational error rates of live and video window counts are unknown, and sampling error rates are ignored. Additionally, two biological processes are unaccounted for: 1) fish that cross the dam during the 8-hours when the window is closed for counting (i.e., nighttime passage) which may result in an underestimate of escapement, and 2) fish that cross the dam multiple times (i.e., re-ascension), and therefore potentially double-counted. Finally, Chinook Salmon and steelhead that migrate through the ladder and past the dam may fallback over the dam (e.g., over spillway, through navigation locks), and later, may or may not re-ascend the fish ladder again [@Boggs2004]. Both fallback with no re-ascension and fallback with re-ascension potentially result in an overestimate of escapement [@Dauble2000]. Previously, it was assumed that nighttime passage rates and fallback/re-ascension rates cancelled each other out resulting in window counts providing an unbiased estimate of escapement (**citation?**)

In this study, we describe a novel method to estimate aggregate and group escapement which incorporates all sources of known uncertainty, we demonstrate the estimation method is essentially unbiased, thus improving effective conservation and management decision making, and show night-time passage and fallback/re-ascension rates are unequal. Our method for estimating escapement, by species, past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of PIT tagged fish in the adult ladder to explicitly model nighttime passage, re-ascension, and error from both window and trap estimates using a state-space approach [@royle2008hierarchical]. To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata. Further, total and weekly estimates are parsed into three origin groups: wild, hatchery, and hatchery no-clip. Estimates of escapement account for fish that migrate through the ladder at night outside of observation hours and account for fish that may ascend the ladder multiple times due to fallback and re-ascension. Our model is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM) package for the statistical software R [@R-Core-Team2020], and is available for downloaded at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM). To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates. The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provide more reasonable estimates of uncertainty. Additionally, the STADEM model framework can be applied elsewhere to estimate aggregate and group escapement, provided a passage barrier containing multiple counting and observational mechanisms exist, such as, a counting window, an adult trap sampling a portion of the run, and PIT tag antenna infrastructure.

# Methods

@@

We used STADEM and three sources of data to estimate Chinook Salmon and steelhead escapement at Lower Granite Dam from 2010-2019. Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of previously PIT tagged fish detected in the adult fish ladder. Following, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; however, similar data could likely be obtained from other fish passage facilities.

@@

Daily counts of adult Chinook Salmon and steelhead passing an observation window located on the Lower Granite fish ladder were the first source of data. Daily counts were made and provided by the US Army Corps of Engineers, and when summed, provide an estimate of the number of fish ascending and passing (i.e., escapement) Lower Granite Dam each season. Window counts were made for each species using video monitoring and direct in-person visual monitoring during daytime hours [@Hatch1994]. Video monitoring occured during the beginning and tail ends of the adult runs (March 1 – March 31 and November 1 – December) for 10 hours per day (0600 – 1600 hours). Direct visual monitoring occured during peak run times (April 1 – October 31) for 16 hours per day (0400 – 2000 hours) [@USACE2015]. During direct visual monitoring, observers recorded each adult (≥ 30cm), by species, passing the window for 50 minutes of each hour of operation. Salmonids under 30cm in length were not identified to species. The sum of the daily 50-minute counts were then multiplied by 1.2 to account for the 10 minutes when fish were not counted. Daytime window counts were not expanded for fish that may have ascended the ladder outside of operational hours (i.e., nighttime) [@USACE2015]. Window counts were accessed through the Columbia Basin Research Data Access in Real Time (DART) website, [www.cbr.washington.edu/dart/query/adult_daily](www.cbr.washington.edu/dart/query/adult_daily), using their window count query. Counts were provided for each day the fish ladder was open to passage. Although window counts were assumed to be a census of every fish passing Lower Granite Dam, corrections were not applied for nighttime passage or re-ascending fish. Further, there was no estimation of daily or seasonal observation or sampling error.

@@

The second source of data came from a sample of fish collected in the adult trap as they migrated past Lower Granite Dam [@Ogden2016b]. The trap, also located within the adult fish ladder and upstream of the observation window, provided biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) for captured adults that allowed decomposition of the escapement into specific groups (e.g., @Camacho2019, @Steinhorst2017). The trap was operational for 24 hours per day, and randomly sampled the daily run by opening four times per hour for a length of time determined by a set daily trapping rate. The trap rate was determined by a committee of collaborating management agencies with a goal of capturing a target number of wild fish, but also balancing fish handling concerns. Trap sample rates were typically 10-25%, but fluctuated throughout the season due to high water temperatures, decreased flows, trap malfunctions and/or closures, fish handling logistics, etc. All captured fish were anesthetized, identified to species, examined for existing marks/tags, and measured for fork length. Additionally, each fish had scale and genetic tissue samples taken; scale samples were used to estimate age [@Wright2015], tissue samples were used to determine sex and for genetic stock identification analysis (e.g., @Hargrove2019). Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag. Final determination of wild, hatchery, and hatchery no-clip origins were assigned using a post-hoc analysis of marks, tags, and genetic information. Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game [@Camacho2019].

### PIT tag Data

The last source of data was observations of PIT tagged adult Chinook Salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder. These observations provide estimates of 1) a trapping rate, 2) the proportion of fish passing during nighttime hours (i.e., outside of window observation hours), and 3) the proportion of fish that ascend the fish ladder multiple times (i.e., re-ascension rate). Detections used in the model include all fish that were previously PIT tagged as juveniles or adults prior to reaching Lower Granite Dam (i.e., does not include newly tagged adults at the dam) and detected at adult detection sites in the dam passage system. PIT tag data was provided through DART and the adult ladder PIT tag query; [http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail](http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail).

A trap rate estimate was derived using mark-recapture methods and PIT tag observations of both Chinook Salmon and steelhead at Lower Granite Dam adult detection sites. The "mark" group included all tags detected in the adult trap and the "capture" group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system. Using a mark-recapture model with differing capture probabilities, we estimated the trap rate on a weekly basis. Those estimates, with associated uncertainties, were then incorporated into the model as informed priors, while the model estimated the true trap rate based on all the data, including trap and window counts. <!--The true trapping rate was estimated as the recorded time of the trap being open does not always reflect the true proportion of fish that are captured in the trap, due to trap malfunctions, sort-by-code fish opening the trap more frequently than expected, or process error (Rick Orme, *per. comm.*). Therefore, we use the mark-recapture estimate of trap rate. -->

The nighttime passage rate was estimated as the count of tags that migrated through the fish ladder during non-window observation hours divided by the total number of tags passing the fish ladder, and was calculated on a weekly basis. The re-ascension rate was the count of tags observed passing the upstream most detection sites in the adult fish ladder (i.e., passing the dam) and later detected re-entering the downstream end of the fish ladder at a later time divided by the total number of tags leaving the fish ladder. Previously, we examined for differences in nighttime passage and re-ascension rates as estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference. Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.

@@

We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage, and fallback/re-ascension rates using a state-space modeling approach [@royle2008hierarchical] implemented in the STADEM package for the R statistical software [@R-Core-Team2020]. We assumed that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) were generated by processes with observation error. In the case of the trap for example, we assumed there was sampling variation and uncertainty around our estimates of the true unknown trap rate. STADEM further accounted for the proportion of fish that ascended the ladder while the counting window was closed (i.e., night), as well as for fish potentially double-counted (or more) after falling back below the dam and later re-ascending the fish ladder. Last, adult sampling data from the trap (wild, hatchery, hatchery no-clip) were used to partition the total escapement estimate by origin (Figure \@ref(fig:stadem-examp-fig). Additional model details can be found in [Appendix A](#append1). The STADEM package is available from the primary author at a [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM), and requires the use of the JAGS software [@R-rjags] for Bayesian inference.

@@

We tested the STADEM model on a variety of simulated data sets. These simulated data sets contained a fixed number of unique adult fish of known origin crossing a dam, from a total of 25 fictional populations with differential run-timing (i.e., date of passage at Lower Granite Dam). Each simulated fish was given a date of ladder ascension, based on its population and the range of run timing for that population. Each fish was also simulated to cross the dam either while the window was open for counting, or not, and was given the chance to be "caught" in the simulated fish trap given the week when it ascended the dam, and the known trap rate that week. Fallback and re-ascension behavior was also simulated, with each fish having the possibility of falling back and re-ascending the ladder up to three times.

Our objective was to examine STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement from the combinations of two separate trapping rates, two fallback/re-ascension and nighttime passage combinations, and three window count error rates; resulting in twelve different scenarios (Table \@ref(tab:lgr-sim-tab)). The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at Lower Granite Dam between 2010-2015. Further details about simulation procedures can be found in [Appendix B](#append2).

@@

Finally, we applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead returning to the Snake River to spawn during 2010 to 2019. Window counts for both species were accessed from DART via functions within STADEM. For Chinook Salmon, a spawn year refers to adults that migrate past the dam prior to August 17 each year and spawn that late summer and fall. For steelhead, the spawn year is defined as steelhead that migrate past Lower Granite Dam starting July 1 the previous year and prior to June 30 of the given year (e.g., spawn year 2017 steelhead migrate past Lower Granite between July 1, 2016 and June 30, 2017). Data from the adult trap was made available by Idaho Department of Fish and Game, and adult PIT tag detection data within the fish passage ladder at Lower Granite Dam was accessed from the PTAGIS regional database ([https://www.ptagis.org/](https://www.ptagis.org/)).

@@

STADEM results were very similar across all scenarios (Figure \@ref(fig:rel-bias-lgd)). Estimates were unbiased, with an average relative bias of 0.2–0.3%. The CV of the estimates average 2.0–3.0%, with coverage probabilities that always exceeded 95%. We calculated RMSD as the square root of the sum of the variance of the estimate and the squared expected bias, which accounts for the size of the uncertainty in the estimator as well as its bias. The RMSD was near 500 for each scenario, representing an estimate within 2% of the true value (Table \@ref(tab:summ-tab)).

@@

We applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead for spawn years 2010–2019. Estimates of total escapement, as well as estimates of wild, hatchery, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab). Coefficients of variation ranged from 3.5-7.5% for wild fish, 3.0-6.2% for hatchery fish, 4.0-8.4% for hatchery no-clip fish and 2.9-5.6% for total unique fish past Lower Granite Dam.

Weekly estimates of total escapement over Lower Granite Dam tracked the window counts and trap estimates (Figure \@ref(fig:jags-fit-fig)). STADEM point estimates were often between estimates based on window counts, and those based on the number of fish caught in the adult trap. However, when very few fish were caught in the trap, or there was more uncertainty about the trap rate that week, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig). That year shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August.

Examining the estimates of nighttime passage and re-ascension rates based on the observed PIT tags crossing over Lower Granite Dam, the two rates did not match up (Figure \@ref(fig:night-reasc-diff-fig)). In particular, there are several weeks when the window counts are quite high, and the rates differ by as much as 10%. Clearly, those two biological processes do not cancel each other out, and thus, employing a model that accounts for both will result in more accurate estimates of escapement.
